Work Fragmentation in Developer Interaction Data

### 2016-01-14 paper_2.tex
Review 2.12 and other minor changes

\subsection{Transformation}
Our goal is to build compact and representative models from each session. In this sense, we used aggregation of events to generate a multivariate time series (MTS). An MTS is a sequence of multivariate observations taken at continuous time intervals coming from a same phenomenon. We build the MTS with edit and navigation variables; the time unit is the minute, and the amplitude is the sum of all the events that occurred in this minute. We selected the minute as unit time because it seemed to be an appropriate and minimal representation of the user interaction in a programming task---obviously a subjective decision. Multivariate time series is a useful way of representing temporal data because they allow us to apply clustering techniques and find recurrent patterns. Additionally, the partition of the data into sessions do not alter the results since we are interested in the values by minute, so having longer sessions do not necessarily increase the number of events per minute.

@@

\subsection{Transformation}
Once the events were classified and sorted, we split the dataset into sessions, defining a session as a period of registered work between interruptions of 8 hours or larger, as we did with Mylyn. We split the data by users to avoid a conflict between the last event of a user and the first one of another. After this, we obtained a total of 28,989 different sessions; but after a filter to keep only those with at least 30 minutes of productive time (as we did with the Mylyn dataset), we ended up with 15,825 useful sessions, in contrast with the 4,284 we had from Mylyn. At this point, we had the same format of sessions with both datasets. Some statistics about the sessions can be seen in the Table \ref{tbl:stats_sessions_udc}, and a histogram of the duration of the sessions on UDC is shown in the Figure \ref{fig:hist_session_duration_udc}, which ignores outliers by setting a threshold of three times the standard deviation.

@@

We measure the effects on productivity using mainly the edit and selection events per minute as indicators. The edit ratio might be also a good indicator, under the assumption that a productive programmer will execute more code editing activities than selection or navigation events. However, this is not always the case.

@@

%First, we analyzed the number of interruptions over the productivity. In this sense, 
We split the data in five groups: The first group contain all the sessions without interruptions (\textit{none}). For the others groups, we have considered four ranges of number of interruptions delimited by their quartiles; the results differ between Mylyn and UDC, existing more interruptions in the former (Table \ref{tbl:quartil_int}). Then, for each group, we display the distribution of events per minute and edit ratio via boxplots. We can see the results with Mylyn in Figure \ref{fig:box_int_events}), and for UDC in Figure \ref{fig:box_int_events_udc}.

@@

\subsection{Effect on the Edit Ratio}
Finally, we want to know the effect the interruptions over the ratio of edits in each session, as this is the often seen as a better indicator of productivity than raw activity, since the programmer spends less time navigating the source code in search of information, and more time actively editing it \cite{KM06}.

@@

As for UDC, the edit ratio behaves similarly in comparison with Mylyn, where the median changes accordingly with the number of interruptions, decreasing when there are more and reaching the maximum when there are none. In addition, the size and form of the boxes indicate more variation in this results and the differences between groups is more subtle.

@@

\begin{table}[ht!]
\caption{Local Effect of Interruption. }
\label{tbl:local_effect}
\centering
\begin{tabular}{p{4cm} | c | c}
effect & pattern in Mylyn & pattern in UDC \\
\hline
\textit{negative}: when the frequency of edit events decreases after the interruption (Cohen's $d < -0.2$)
	& \includegraphics[valign=m,scale=0.3]{figures/neg_ints}
	& \includegraphics[valign=m,scale=0.25]{figures/udc_neg_ints_2}  \\
\textit{positive}: when the frequency of edit events increases after the interruption (Cohen's $d > 0.2$)
	& \includegraphics[valign=m,scale=0.3]{figures/pos_ints}
	& \includegraphics[valign=m,scale=0.25]{figures/udc_pos_ints_2} \\
\textit{neutral}: when there is no well defined effect before or after the interruption ($abs(d) <= 0.2$)
	& \includegraphics[valign=m,scale=0.3]{figures/neu_ints} 
	& \includegraphics[valign=m,scale=0.25]{figures/udc_neu_ints_2} 
\end{tabular}
\end{table}

The results in UDC indicate that after a positive interruption both metrics tend to increase at a high rhythm until reaching the global average value. The editions per minute recover quickly, approximately 5 minutes after the interruption, but the selections per minute take a little longer to reach the global average and this metric tend to be below the average for an extended period. It is important to notice that before the interruption happens the editions per minute metric is below the average for approximately 16 minutes prior to the interruption.

@@

From those valid interruptions, we split the time after the interruption occurred into two phases:

@@

The duration of the peak varies among the also called working spheres, as mentioned by Gonzalez and Mark \cite{MGH05}. On average, the peak lasts for 12 minutes, while the maximum registered is of 145 minutes. The distribution of duration of the peak is reported in Table \ref{tbl:dist_peaks}.

\begin{table}[ht!]
	\small
	\renewcommand{\arraystretch}{1.3}
	\caption{Quartiles of the distribution of the duration of the peak phase.}
	\label{tbl:dist_peaks}
	\centering
	\begin{tabular}{l | p{0.6cm} | p{0.6cm} | p{0.6cm}} 
		& $25\%$ & $50\%$ & $75\%$ \\  
		\hline 
		quartile & 3.38 & 8.05 & 16.31  \\   
	\end{tabular}
\end{table}

@@

To better understand the reasons of the difference between positive and negative interruptions, we did an analysis of the common events during recovery time, depending on the nature of the interruption. For this we created a new classification of events on a lower level that takes into consideration what kind of activity is performed by the programmer. The new classification contains the categories described in the Table \ref{tbl:new_events_udc} and the frequency per minute of the more frequent event on that category. In a similar manner to the first classification, we followed an iterative process involving two of the authors to classify the events in these categories.

@@

We calculated the weighted average of every event based on the frequency of execution during the first phase of recovery time after a positive or negative interruption, and the percentage of users that make use of that event (so that events that are used by very few users do not distort the results). The weighted average was calculated as $\bar{x} = \sum_{i=1}^{n}w_ix_i / \sum_{i=1}^{n}w_i$, where $x_i$ is de frequency of execution of the event $i$ and $w_i$ is the percentage of users that execute the event. This balances the average of execution for events used by a small number of users. For example, the event \textit{copyLineDown} is more frequent than the \textit{rename} event, but the former is only used by 2.4\% of the users and the latest by 25.2\% of them.

To get a baseline, the same frequency calculation was made for the entire dataset, without distinction of the location of events in the session. We then compared the weighted average changes to have evidence of what kind of events are more often executed after a positive or a negative interruption. Table \ref{tbl:stats_events} shows the results.

@@

After a positive interruption we can see that the more common activities are text edition, refactoring, file management and, to a lesser extent, navigation through the text. On the other hand the navigation around classes, find and replace, searches, debug, clean and build are actions more common after a negative. The navigation in the code and control version activities are more common after the Recovery Time.

We can conclude that after a positive interruption there are more assertive actions with a clear intention like refactoring a class or saving all changes; activities related to program edition. After a negative interruption there are more uncertainty and doubt judging by the activities performed like debugging, searching for text or navigating around classes; activities related to program comprehension. These categories of events agree with our previous hypothesis that positive interruptions are related to information seeking \cite{PR11, LVD06}, while negative interruptions are related to actual task switches where context needs to be rebuilt afterwards.

@@

\paragraph{Comparison with observational studies}
Our results and theoretical foundations are based on observational studies performed on information workers' development environments \cite{MGH05, IH07, LVD06, PD10}. These findings helped us understand the response to work fragmentation according to the recorded user activity, and it was possible to made a relation between the user's data and the phenomena that occur on the working places describe on the studies.

Observing the user actvivity throughout a working session might be the best way to understand the effect of events that occur on the working place, like work fragmentation. However, implementing a study of this kind can be very expensive, for the time and material need. For example, to have the same amount of information we have with UDC it would require to monitor the user activity for a year or two and have a record of all the activities performed, like the number and duration of the interruptions. Moreover, this studies require on-site studies, exhaustive monitoring and interviews, which in some cases could be expensive or difficult to perform.

@@

The best might be to find a middle ground between observational activities and analysis of usage data. It is helpful to enhance the results using information provided directly by the participants via interviews, like in \cite{FMH07}.

>>>>

\subsection{Transformation}
Our goal is to build compact and representative models from each session. In this sense, we used aggregation of events to generate a multivariate time series (MTS). An MTS is a sequence of multivariate observations taken at continuous time intervals coming from a same phenomenon. We build the MTS with edit and navigation variables; the time unit is the minute, and every observation is the sum of all the events that occurred in this minute. We selected the minute as unit time because it seemed to be an appropriate and minimal representation of the user interaction in a programming task---obviously a subjective decision.

Multivariate time series is a useful way of representing temporal data because they allow us to apply clustering techniques and find recurrent patterns. Additionally, the partition of the data into sessions do not alter the results since we are interested in the values by minute, so having longer sessions do not necessarily increase the number of events per minute.

@@

\subsection{Transformation}
Once the events were classified and sorted, we split the dataset into sessions, defining a session as a period of registered work between interruptions of 8 hours or larger, as we did with Mylyn. We split the data by users to avoid a conflict between the last event of a user and the first one of another. After this, we obtained a total of 28,989 different sessions; but after a filter to keep only those with at least 30 minutes of productive time (as we did with the Mylyn dataset), we ended up with 15,825 useful sessions, in contrast with the 4,284 we had from Mylyn. On average, every session contains 154 edition and selection events, and this number is positively correlated (0.81) with the size of the session. This is an expected result, as having more time consequently increases the number of events in the session. As mentioned before, the number of events in a session do not alter the edit or selection events per minute.

At this point, we had the same format of sessions with both datasets. Some statistics about the sessions can be seen in the Table \ref{tbl:stats_sessions_udc}, and a histogram of the duration of the sessions on UDC is shown in the Figure \ref{fig:hist_session_duration_udc}, which ignores outliers by setting a threshold of three times the standard deviation.

@@

We measure the effects on productivity using mainly the edit and selection events per minute as indicators. The edit ratio might be also a good indicator, under the assumption that a productive programmer will execute more editing activities than selection or navigation events. However, this is not always the case, given that depending on the role of the worker, selection events can be equivalent to the usage of a designing or reporting tool.

@@

%First, we analyzed the number of interruptions over the productivity. In this sense, 
We split the data in five groups: The first group contains all the sessions without interruptions (\textit{none}). For the others groups, we have considered four ranges of number of interruptions delimited by their quartiles; the results differ between Mylyn and UDC, existing more interruptions per session in the former (Table \ref{tbl:quartil_int}). Then, for each group, we display the distribution of events per minute and edit ratio via boxplots. We can see the results with Mylyn in Figure \ref{fig:box_int_events}), and for UDC in Figure \ref{fig:box_int_events_udc}.

@@

\subsection{Effect on the Edit Ratio}
Finally, we want to know the effect of the interruptions over the ratio of edits in each session, as this is the often seen as a better indicator of productivity than raw activity, since the programmer spends less time navigating the source code in search of information, and more time actively editing it \cite{KM06}.

@@

As for UDC, the edit ratio behaves similarly in comparison with Mylyn, where the median changes accordingly with the number of interruptions, decreasing when there are more and reaching the maximum when there are none. In addition, the size and form of the boxes indicate more variation in this results and the differences between groups is more subtle, judging by the small changes in the mean values.

@@

\begin{table}[ht!]
\caption{Local Effect of Interruption. }
\label{tbl:local_effect}
\centering
\begin{tabular}{p{5cm} | c | c}
effect & pattern in Mylyn & pattern in UDC \\
\hline
\textit{negative}: when the frequency of edit events decreases after the interruption (Cohen's $d < -0.2$)
	& \includegraphics[valign=m,scale=0.3]{figures/neg_ints}
	& \includegraphics[valign=m,scale=0.24]{figures/udc_neg_ints_2}  \\
\textit{positive}: when the frequency of edit events increases after the interruption (Cohen's $d > 0.2$)
	& \includegraphics[valign=m,scale=0.3]{figures/pos_ints}
	& \includegraphics[valign=m,scale=0.24]{figures/udc_pos_ints_2} \\
\textit{neutral}: when there is no well defined effect before or after the interruption ($abs(d) <= 0.2$)
	& \includegraphics[valign=m,scale=0.3]{figures/neu_ints} 
	& \includegraphics[valign=m,scale=0.24]{figures/udc_neu_ints_2} 
\end{tabular}
\end{table}

The results in UDC indicate that after a positive interruption, both metrics tend to increase at a high rhythm until reaching the global average value. The edits per minute recover quickly (approximately 5 minutes after the interruption), but the selections per minute take a little longer to reach the global average, and this metric tend to be below the average for an extended period. It is important to notice that before the interruption happens the editions per minute metric is below the average for approximately 16 minutes prior to the interruption.

@@

From those valid interruptions, we split the time after the interruption into two phases:

@@

The duration of the peak varies among the also called working spheres, as mentioned by Gonzalez and Mark \cite{MGH05}. On average, the peak lasts for 12 minutes, while the maximum registered is of 145 minutes. The distribution of duration of the peaks is reported in Table \ref{tbl:dist_peaks}.

\begin{table}[ht!]
	\small
	\renewcommand{\arraystretch}{1.3}
	\caption{Quartiles of the distribution of the duration of the peak phase.}
	\label{tbl:dist_peaks}
	\centering
	\begin{tabular}{l |c | c |c} 
		quartile & $25\%$ & $50\%$ & $75\%$ \\  
		\hline 
		 value (min) & 3.38 & 8.05  & 16.31   \\   
	\end{tabular}
\end{table}

@@

To better understand the reasons of the difference between positive and negative interruptions, we did an analysis of the common events during recovery time, depending on the nature of the interruption. For this we created a new classification of events on a lower level that takes into consideration what kind of activity is performed by the programmer. The new classification contains the categories described in the Table \ref{tbl:new_events_udc}. This table also contains the frequency per minute of the more frequent event on that category. In a similar manner to the first classification, we followed an iterative process involving two of the authors to classify the events in these categories.

@@

We calculated the weighted average of every event based on the frequency of execution during the first phase of recovery time after a positive or negative interruption, and the percentage of users that make use of that event (so that events that are used by very few users do not distort the results). The weighted average was calculated as: $$\bar{x} = \frac{\sum_{i=1}^{n}w_ix_i}{\sum_{i=1}^{n}w_i} $$where $x_i$ is de frequency of execution of the event $i$ and $w_i$ is the percentage of users that uses the event. This balances the average of execution for events used by a small number of users. For example, the event \textit{copyLineDown} is more frequent than the \textit{rename} event, but the former is only used by 2.4\% of the users and the latest by 25.2\% of them.

To get a baseline, the same frequency calculation was made for the entire dataset, without distinction of the location of events in the session. We then compared the weighted average changes to have evidence of what kind of events are more often executed after a positive or a negative interruption. The Table \ref{tbl:stats_events} shows the results.

@@

After a positive interruption, we can see that the more common activities are text edition, refactoring, file management and, to a lesser extent, navigation around perspectives. On the other hand the navigation around classes, find and replace, searches, debug and clean and build are actions more common after a negative one. The navigation in the code and control version activities are more common after the recovery time, having a lower frequency during a positive o negative interruption.

We can conclude that after a positive interruption there are more assertive actions with a clear intention like refactoring a class or saving all changes; these activities are related to program edition. After a negative interruption, there are more uncertainty and doubt judging by the activities performed like debugging, searching for text or navigating around classes; these are activities related to program comprehension. These categories of events agree with our previous hypothesis that positive interruptions are related to information seeking \cite{PR11, LVD06}, while negative interruptions are related to actual task switches where context needs to be rebuilt afterwards.

@@

\paragraph{Comparison with observational studies}
Our theoretical foundations are based on observational studies performed on information workers' development environments \cite{MGH05, IH07, LVD06, PD10}. These findings helped us understand the response to work fragmentation according to the recorded user activity, and it was possible to made a relation between the user's data and the phenomena that occur on the working places describe on the studies.

Observing the user actvivity throughout a working session might be the best way to understand the effect of events that occur on the working place, like work fragmentation. However, implementing a study of this kind can be very expensive, for the time and material need. For example, to have the same amount of information we have with UDC it would require to monitor the user activity for a year or two and have a record of all the activities performed, like the number and duration of the interruptions. Moreover, this kind of research require on-site studies, exhaustive monitoring and interviews, which in some cases could be expensive or difficult to perform.

@@

The best might be to find a middle ground between observational activities and analysis of usage data. It is helpful to enhance the results using information provided directly by the participants via interviews, like the work done by Fritz et al. \cite{FMH07}.
